Large Linear Multi-output Gaussian Process Learning for Time Series

### 2017-05-07 paper/paper.tex
icml -> nips paper conversion

\documentclass{article}
\usepackage{times}
\usepackage{graphicx}
\usepackage{subfigure} 
\usepackage{natbib}
\usepackage{algorithm}
\usepackage[nice]{nicefrac}
\usepackage{algorithmic}
\usepackage{hyperref}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{icml2017}
%\usepackage[accepted]{icml2017}

\def\bsa{{\boldsymbol\alpha}}
\def\bse{{\boldsymbol\epsilon}}
\def\bsk{{\boldsymbol\kappa}}
\def\bsm{{\boldsymbol\mu}}
\def\bsl{{\boldsymbol\lambda}}
\def\bsph{{\boldsymbol\phi}}
\def\bsth{{\boldsymbol\theta}}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{enumerate}
\usepackage{fullpage}
\usepackage{bbm}
\usepackage{caption}
\usepackage{mathrsfs}

\newcommand{\pa}[1]{ \left({#1}\right) }

\def \N {\mathbb{N}}
\def \Nbr {\mathcal{N}}
\def \mcL {\mathcal{L}}
\def \Q {\mathbb{Q}}
\def \F {\mathbb{F}}
\def \then {\implies &}
\def \oif {\Longleftrightarrow &\,}
\def \given {\text{Given }&}
\def \assume {\text{Assume }&}
\def \thfr {\therefore &\enskip}
\def \bij {\leftrightarrow}
\def \inj {\rightarrowtail}
\def \sur {\twoheadedrightarrow}
\def \Z {\mathbb{Z}}
\def \R {\mathbb{R}}
\def \C {\mathbb{C}}
\def \T {\mathbb{T}}
\def \iff {\Longleftrightarrow}
\def \kron {\boldsymbol\delta}
\def \indicator {\mathbbm{1}}

\def\Tx{\textbf{x}}
\def\TX{\textbf{X}}
\def\TU{\textbf{U}}
\def\Ty{\textbf{y}}
\def\tbr{\textbf{r}}
\def\Tz{\textbf{z}}
\def\quotient{\mathclose{}/\mathopen{}}
\def\Tf{\textbf{f}}
\def\Th{\textbf{h}}
\def\Tg{\textbf{g}}
\def\sumn{\sum_{n=0}^\infty}
\def\limn{\lim_{n\rightarrow\infty}}
\def\prodn{\prod_{n=0}^\infty}

\DeclareMathOperator\adj{adj}
\DeclareMathOperator\cov{cov}
\DeclareMathOperator\Unif{Unif}
\DeclareMathOperator\blockdiag{blockdiag}
\DeclareMathOperator\Pois{Pois}
\DeclareMathOperator\mathProb{\mathbb{P}}
\def\P{\mathProb} % need to overwrite stupid paragraph symbol
\DeclareMathOperator\mathExp{\mathbb{E}}
\def \E {\mathExp}

\newcommand{\stc}[1]{\widetilde{#1}}   
\newcommand{\set}[2]{ \left\{ #1 \,\middle|\, #2 \right\} }
\newcommand{\shift}[1]{&\quad & \text{#1}\\}
\newcommand{\lem}[1]{\text{\textbf{L.\ref{#1}}}}
\newcommand{\card}[1]{\left\vert{#1}\right\vert}
\newcommand{\Ps}[1]{\mathcal{P}\left({ #1 }\right)}
\newcommand{\colv}[1]{\begin{pmatrix} #1 \end{pmatrix}}
\newcommand{\mat}[1]{\begin{pmatrix} #1 \end{pmatrix}}
\newcommand{\detmat}[1]{\begin{vmatrix} #1 \end{vmatrix}}
\newcommand{\spanb}[1]{\text{span}\{ #1 \}}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\Inner}[1]{\langle #1 \rangle}
\newcommand{\Innercpy}[1]{\langle #1, #1 \rangle}
\newcommand{\norm}[1]{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$

\def\eqd{\mathrel{\overset{\Delta}{=}}}

\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\vcdim}{VC-dim}
\DeclareMathOperator*{\Err}{\text{err}}
\DeclareMathOperator*{\ErrE}{\mathbb{E}}
\DeclareMathOperator{\Tr}{tr}
\DeclareMathOperator{\Dim}{dim}
\DeclareMathOperator{\Rank}{rank}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\proj}{proj}
\DeclareMathOperator{\Ker}{ker}
\DeclareMathOperator{\Diam}{diam}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clo}{clo}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\MyRe}{Re}
\DeclareMathOperator{\MyIm}{Im}
\DeclareMathOperator{\Supp}{supp}
\DeclareMathOperator{\Var}{Var}

% probability stuff
\def \sa {{$\sigma$-algebra}}
\def\OR{{\overline{\R}}}
\def\OX{{\overline{X}}}

\def \mcX {\mathcal{X}}
\def \mcY {\mathcal{Y}}
\def \mcH {\mathcal{H}}
\def \mcD {\mathcal{D}}
\def \mcA {\mathcal{A}}
\def \mcK {\mathcal{K}}
\def\mcG{{\mathcal{G}}}
\def\mcH{{\mathcal{H}}}
\def\mcF{{\mathcal{F}}}
\def\mcB{{\mathcal{B}}}
\def\mcE{{\mathcal{E}}}
\def\mcP{{\mathcal{P}}}
\def\mcM{{\mathcal{M}}}
\def\mcC{{\mathcal{C}}}
\def\mcT{{\mathcal{T}}}

% \icmltitlerunning{Submission and Formatting Instructions for ICML 2017}

@@

\twocolumn[
\icmltitle{Large Linear Multi-output Gaussian Process Learning}
%\icmlsetsymbol{equal}{*}
\begin{icmlauthorlist}
  \icmlauthor{Vladimir Feinberg}{pu}
  \icmlauthor{Li-Fang Cheng}{pu}
  \icmlauthor{Barbara E. Engelhardt}{pu}
  \icmlauthor{Kai Li}{pu}
\end{icmlauthorlist}
\icmlaffiliation{pu}{Princeton University, Princeton, USA}
\icmlcorrespondingauthor{Vladimir Feinberg}{vyf@princeton.edu}

\icmlkeywords{multi-task, multi-output, Gaussian process, large scale}

\vskip 0.3in
]
\printAffiliationsAndNotice{}
%\printAffiliationsAndNotice{\icmlEqualContribution}

\begin{abstract}
Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization \cite{alvarez2012kernels}. Such models can learn correlations across the multiple outputs, such as temperature data from disparate regions, which can then be exploited to share knowledge across the multiple outputs. While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large $n$, the total number of observations across all outputs. In this work, we describe a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs. We show results of our approach applied to real time series data, and compare these results against related methods. Finally, we discuss extensions of our approach to multidimensional inputs.
\end{abstract}

\section{Introduction}
\label{introduction}

@@

Such an extension is especially important in time series data, where multi-output GPs have been applied to the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.

Exact GP inference comes at a cost: to compute the gradients for kernel hyperparameters, an exact computation requires an $O(n^3)$ decomposition of an $O(n^2)$ matrix in memory for $n$ samples \cite{gpml}. Each hyperparameter's scalar partial derivative then requires another $O(n^2)$ computation. While these operations may be performed in parallel, the memory use and runtime is prohibitive for problems with a large number of outputs, since the number of hyperparameters grows accordingly. For this reason, an accurate approximate approach has been an important goal of machine learning research.

For a single output, variational approaches that work on $m\ll n$ inducing points, or locations at which the true kernel is sampled, letting users trade off accuracy for efficiency \cite{hensman2013gaussian}.

Nguyen and Bonilla observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once \yrcite{nguyen2014collaborative}. We analogously exploit sharing in the covariance structure.

@@

Our approach makes the following contributions to approximate inference in multi-output GPs. First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid. Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}. Both of these structures coordinate for fast matrix-vector multiplication $K\Ty$ with the covariance matrix $K$.

When inputs do not lie on a uniform grid (i.e., for time series, the length of time between observations differs), we apply a technique proposed in Massively Scalable Gaussian Processes (MSGP) \cite{msgp}. For single-output kernels, MSGP leverages the structured kernel interpolation (SKI) framework to make inputs act as if they are on a grid \cite{kiss-gp}. In a single-output case, this opens up the possibility for a matrix-free GP, avoiding construction of the explicit Gram covariance matrix. Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI. While direct application of SKI is not possible, we can make an extension that makes matrix-free LMC kernel learning viable even on non-grid inputs.

Our paper is organized as follows. In Section~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs. Section~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs. Section~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion. Then, in Section~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Section~\ref{conclusion}.

\section{Background}
\label{sec:background}
\subsection{General Gaussian processes}

@@

Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by marginalization \cite{gpml}:
\begin{equation*}
y_*|\Ty\sim N\pa{K_{*, {X}}K^{-1}_{{X}, {X}}\Ty, K_{*, *}-K_{*, {X}}K^{-1}_{{X}, {X}}K_{{X}, *}}.
\end{equation*}
Accuracy is often sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
\begin{align}
  \mcL(\bsth)&=\log p(\Tf_{X}|X,\bsth)\label{datall}\\
             &=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.
\end{align}
As such, the heart of GP learning lies in optimization of $\mcL$. Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
\begin{align}
\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}}.\label{llgradient}
\end{align}

@@

@@

The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$-th block $A_{ij}B$. We can represent it by keeping representations of $A$ and $B$ separately, rather than the product. Then, the corresponding MVMs can be computed in time $O(a\text{MVM}(B)+b\text{MVM}(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM. For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs by a $(1/P)$-th power \cite{gilboa2015scaling}.

Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_j$, yielding an $O(n)$ representation. Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs. Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that:
\[
T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}
\]
Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs  $C\mat{\Tx\\\textbf{0}}=\mat{T\Tx\\\textbf{0}}$ in $O(n\log n)$ time. This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \cite{cunningham2008fast}.

\section{Related Work}
\label{sec:related-work}
\subsection{Approximate inference methods}

To cope with the lack of tractable GP inference methods, inducing point approaches create a tractable model to use instead of the classic GP. Such approaches fix or estimate inducing points $\TU$ and claim that the data $f_X$ is conditionally deterministic (deterministic training conditional, or DTC), independent (fully independent training conditional, or FITC), or partially independent (partially independent training conditional, or PITC) given RVs $f_{\TU}$  \cite{quinonero2005unifying}. These approaches are agnostic to kernel stationarity, and their quality can be improved by increasing the number of inducing points at the cost of a longer runtime. Setting the inducing points $\TU=\TX$ recovers the original exact GP.

In the context of LMC, a framework for variational DTC, DTCVAR, provided a matrix-free method where each iteration has complexity $O(n(Dm)^2)$, where $m$ is the average number of inducing points per output, $\card{\TU}/D$ \cite{alvarez2010efficient}.

By sharing the $m$ inducing points across all outputs, Collaborative Multi-output Gaussian Processes (COGP) improves runtime per iteration to $O(m^3)$. Moreover, COGP only requires $O(Dnm)$ memory \cite{nguyen2014collaborative}. To ensure that $O(m^3)$ is computationally tractable for COGP, $m$ must be sufficiently small. COGP makes large multi-output GP estimation feasible through a variational approximation, the evidence lower bound, to the log likelihood.

@@

\section{Matrix-free LMC Learning} \label{sec:matrix-free}

We propose a linear model of coregionalization (LMC) method based on recent structure-based optimizations for GP estimation instead of variational approaches. Critically, the accuracy of the method need not be reduced by keeping $m$ low because its reliance on structure allows better asymptotic performance.
For simplicity, our work focuses on multi-dimensional outputs, one-dimensional inputs, and Gaussian likelihoods.

Although Eq.~\ref{datall} does not have a hyperprior $\log p(\bsth)$ term, it can easily be augmented with one.

\subsection{Matrix-free GP Learning}

Gibbs and MacKay describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix \yrcite{gibbs1996cient}. In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG). Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
\begin{align}
  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}
\end{align}
For this approximation, the number of samples need not be large, and the estimate improves as the size of $K_{|\bsth}$ increases.
As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$.

We depart from Gibbs and MacKay in selecting MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs. MINRES handles numerically semidefinite matrices with more grace \cite{fong2012cg}. This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations (Fig.~\ref{fx2007-iterations}).

\begin{figure}[!ht]
\vskip 0.1in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{iterations.eps}}
\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Section~\ref{fx2007-results}. The iteration cutoff is the number of training points $n$ in the dataset.
}
\label{fx2007-iterations}
\end{center}
\vskip -0.1in
\end{figure}

If we sample $N_t$ instances of $\tbr$, every iteration needs to perform $N_t+1$ inversions including $\Ty$, if we reuse the same samples $K_{|\bsth}^{-1}\tbr$ between derivatives $\partial_{\theta_j}$. These inversions are done in parallel. Then, since $\partial_{\theta_j}K_{|\bsth}$ is both structured and smaller than $K_{|\bsth}$, each of the terms in the approximation Eq.~\ref{eq:trace} require $\tilde{O}(n)$ time to compute.

Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine \cite{zeiler2012adadelta}. We considered several stochastic approximations for finding $\log\abs{K_{|\bsth}}$~\cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.

In other words, all the work is done in the inversions $K_{|\bsth}^{-1}\tbr,K_{|\bsth}^{-1}\Ty$, which in turn rely on the MVM operation, discussed in Section~\ref{fast-mvm}. Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation, which need not be dense.

@@

When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$.
\begin{align}
  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}
\end{align}
The SLFM and COGP models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$. Moreover, we include $D$ additional kernels for the independent GP components, which can be incorporated using kernels $K_d$ where $A_d=0$ and $\bsk_d$ as the $d$th standard basis vector of $\R^D$ for $d\in[D]$. This shows a reduction from SLFM to LMC.

>>>>

\documentclass{article}

\PassOptionsToPackage{numbers, compress}{natbib}
% Camera ready: \usepackage[final]{nips_2017}
\usepackage{nips_2017}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{booktabs}
\usepackage{subfigure}

\input{defs}

\title{Large Linear Multi-output Gaussian Process Learning for Time Series}

\author{
  Vladimir Feinberg, Li-Fang Cheng, {Barbara Engelhardt}, {Kai Li}\\
  Department of Computer Science\\
  Princeton University\\
  Princeton, NJ 08544 \\
  \texttt{\{vyf,lifangc,bee,li\}@princeton.edu} \\
}

@@

\maketitle

\begin{abstract}
Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization \cite{alvarez2012kernels}. Such models can learn correlations across the multiple outputs, which can then be exploited to share knowledge across the multiple outputs. For instance, temperature data from disparate regions over time can contribute to a predictive weather model that is more accurate than the same model applied to a single region. While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large numbers of observations across all outputs. In this work, we propose a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs. Our approach learns GP kernel hyperparameters at an asymptotically faster rate than the current state of the art. When applied to real time series data, we find this theoretical improvement is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy. Finally, we discuss extensions of our approach to multidimensional inputs.
\end{abstract}

% NOTES
% 8 pages of content
% \citet for inline citations
% footnotes after punctuation
% no vertical rules in tables
% \subsubsection*{Acknowledgments} only in final paper

\section{Introduction}\label{introduction}

@@

Multiple-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}. Efficient model learning would enable researchers to quickly explore large spaces of model configurations to find an appropriate one for their task.

Exact GP inference is infeasible in this scenario since it requires maintaining pairwise covariance between all inputs in a large matrix and performing inversions with that matrix \cite{williams1996gaussian}. As the number of hyperparameters grows, so do the number of matrix operations (Tab.~\ref{asymp}). For this reason, an accurate approximate approach has been an important goal of machine learning research.

@@

Our approach makes the following contributions to approximate inference in multi-output GPs, scoped to time series input. First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid. Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}. Both of these structures coordinate for fast matrix-vector multiplication $K\Ty$ with the covariance matrix $K$.

When inputs do not lie on a uniform grid (for time series, the length of time between observations differs), we demonstrate how to leverage structured kernel interpolation (SKI) in the multi-output setting, where SKI, which requires stationary kernels, does not naturally apply \cite{kiss-gp}. Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI.

For low-dimensional inputs, the above contributions offers an asymptotic and practical runtime improvement in hyperparameter learning while also expanding feasible kernel typs to arbitrary differntiable LMC kernels, with improvement taken to be relative to existing multi-GP methods (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.

Our paper is organized as follows. In Sec.~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs. Sec.~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs. Sec.~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion. Then, in Sec.~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Sec.~\ref{conclusion}.

\section{Background}
\label{sec:background}

\subsection{General Gaussian processes}

@@

Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty$ \cite{williams1996gaussian}.
Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
\begin{align}
  \mcL(\bsth)=\log p(\Tf_{X}|X,\bsth)=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.\label{datall}
\end{align}
As such, the heart of GP learning lies in optimization of $\mcL$. Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
\begin{align}
\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}}.\label{llgradient}
\end{align}

@@

Thus, the LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between to inputs for two different outputs. The contribution of the $q$-th kernel $k_q$ to the covariance between the $i$-th and $j$-th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$. By choosing $B_q$ to have rank $R_q=D$, so the contribution between those two outputs is learned independently of all other contributions, the entries of $B_q$. By lowering the rank $R_q$, researchers applying the LMC model can specify that the contributions of the kernels for each output pair are determined by lower-rank latent processes, with rank 0 indicating no interaction.

@@

The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$-th block $A_{ij}B$. We can represent it by keeping representations of $A$ and $B$ separately, rather than the product. Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM. For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs by a $(1/P)$-th power \cite{gilboa2015scaling}.

Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_j$, yielding an $O(n)$ representation. Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs. Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that:
\[
T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}
\]
Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs $C\mat{\Tx & \textbf{0}}^\top=\mat{T\Tx &\textbf{0}}^\top$ in $O(n\log n)$ time. This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \cite{cunningham2008fast}.

\section{Related work}
\label{sec:related-work}
\subsection{Approximate inference methods}

To cope with the lack of tractable GP inference methods, inducing point approaches create a tractable model to use instead of the classic GP. Such approaches fix or estimate inducing points $\TU$ and claim that the data $f_X$ is conditionally deterministic (deterministic training conditional, or DTC), independent (fully independent training conditional, or FITC), or partially independent (partially independent training conditional, or PITC) given RVs $f_{\TU}$  \cite{quinonero2005unifying}. These approaches are agnostic to kernel stationarity, and their quality can be improved by increasing the number of inducing points at the cost of a longer runtime. Setting the inducing points $\TU=\TX$ recovers the original exact GP. Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.

\citet{nguyen2014collaborative} observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once. COGP further uses a variational approximation, the evidence lower bound, to the log likelihood. COGP supports a subset of LMC kernels, namely those that match the SLFM model \cite{seeger2005semiparametric}. In an LMC representation (Eq.~\ref{lmcpointwise}), these models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$. Moreover, SLFM and COGP models add in an independent GP to each output, represented in LMC as additional kernels $\idx{k_d}{d=1}{D}$, where $A_d=0$ and $\bsk_d=\Te_d\in\R^D$.

@@

\section{Matrix-free LMC learning} \label{sec:matrix-free}

We propose a linear model of coregionalization (LMC) method based on recent structure-based optimizations for GP estimation instead of variational approaches. Critically, the accuracy of the method need not be reduced by keeping the number of interpolation points $m$ low because its reliance on structure allows better asymptotic performance.
For simplicity, our work focuses on multi-dimensional outputs, one-dimensional inputs, and Gaussian likelihoods.

\subsection{Matrix-free GP learning}

For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$. Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$. Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}. We considered several stochastic approximations for finding the pathological term $\log\abs{K_{|\bsth}}$ in Eq.~\ref{datall} \cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.

\subsection{Gradient construction}

\citet{gibbs1996cient} describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix. In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG). Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
\begin{align}
  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}
\end{align}
For this approximation, the number of samples need not be large, and the estimate improves as the size of $K_{|\bsth}$ increases.
As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number of $N$ samples from $\tbr$.

We depart from Gibbs and MacKay in two ways, yielding Algorithm~\ref{alg:grad}. First, we do not construct $K_{|\bsth}$, but a low-memory representation $\tilde{K}_{|\bsth}$, described in Sec.~\ref{fast-mvm}. Second, we select MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs. MINRES handles numerically semidefinite matrices with more grace \cite{fong2012cg}. This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations because of increases in $\kappa_2$, the spectral condition number of $K$ (Fig.~\ref{fx2007-iterations}).

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\columnwidth]{iterations.eps}
\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}. The iteration cutoff is the number of training points $n$ in the dataset.
}
\label{fx2007-iterations}
\end{figure}

Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K}_{|\bsth})\sqrt{\kappa_2})$ \cite{raykar2007fast}. This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K_{|\bsth}$ for each $j$ at least as fast as $\mvm({\tilde{K}_{|\bsth}})$. Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K_{|\bsth}$.

\begin{algorithm}[!ht]
  \caption{Compute an approximation of $\nabla L$. Assume \textsc{minres} is the inversion routine. We also assume we have access to linear operators $D_{\theta_j}$, representing matrices $\partial_{\theta_j}\tilde{K}_{|\bsth}$.} \label{alg:grad}
\begin{algorithmic}[1]
  \Procedure{LLGP}{$\tilde{K}_{|\bsth}$, $\Ty$, $N$, $\{D_{\theta_j}\}$}
  \State $R\gets\idx{\tbr_i}{i=1}{N}$, sampling $\tbr\sim\Unif\{\pm 1\}$.
\For{$\Tz$ in $\{\Ty\}\cup R$, in parallel}
\State $K^{-1}\Tz\gets\textsc{minres}(\tilde{K}_{|\bsth},\Tz)$.
\EndFor
\State $g\gets \bsz$
\For{$\theta_j$ in $\bsth$}\Comment{Compute $\partial_{\theta_j}\mcL$}
\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot D_{\theta_j}(\tbr_i)$\Comment{$t$ approximates the trace term of Eq.~\ref{eq:trace}}
\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}_{|\bsth}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
\EndFor
\State \Return $g$ 
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The bottleneck of Algorithm~\ref{alg:grad} is the iterative MVM operations in $\textsc{minres}$. Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation $\tilde{K}_{|\bsth}$, which need not be dense, so long as it can reconstruct multiplication with any vector to arbitrary, fixed precision.

When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$, the shared interpolating points for all outputs:
\begin{align}
  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}.
\end{align}

Importantly, the Kronecker structure of Eq.~\ref{lmcgrid} lets us re-use the same grid $K_q$ in a computational sense across the different outputs. Recalling our SKI extension to multiple outputs (Sec.~\ref{ski-section}), we build a corresponding approximation for the differentiable part of our kernel:
\begin{align}
  K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top+\bse.\label{lmcski}
\end{align}
We cannot fold $\bse$ into the interpolated term $K_{\TU,\TU}$ since it does not correspond to a differentiable kernel, so the SKI approximation fails. But this fact does not prevent efficient representation or multiplication since the matrix is diagonal. Then, the MVM operation $K_{\TX,\TX}\Tz$ can be approximated by $WK_{\TU,\TU}W^\top\Tz+\bse\Tz$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ require $O(n)$ space and time.

We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$.
